---
title: "Predicting F1 Pitstops"
author: "Niklas Pettersson-Fox"
date: "2026-05-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Inital set up of data and cleaning.

```{r dataset,  echo = FALSE}
set.seed(123)
train <- read.csv("train.csv")
```


```{r data1, echo = FALSE}
train$Compound <- as.factor(train$Compound)
train$Race <- as.factor(train$Race)
train$Year <- as.factor(train$Year)
train$PitStop <- as.factor(train$PitStop)
train$PitNextLap <- as.factor(train$PitNextLap)
summary(train)
```

As depicted in the summary we can see some unusual values that need to be taken care of. Lap time of 2507 needs to be removed as in F1, lap times are usually in the range of 1-2 minutes or 60-120 seconds hence we will remove any lap times larger than 10 minutes as this will give some leeway for laps where it was raining. This will take care of the Laptime delta  which has a few extreme cases. My guess is that this lap is due to some racing incident, leaving the driver unable to finish the lap whilst the clock was still running. Let's have a look.

```{r laptime, echo = FALSE}
lap_toolarge_i <- which(train$LapTime..s. > 600)
lap_tooLarge_train <- train[lap_toolarge_i,]
train <- train[-lap_toolarge_i,]
```

## Visualisation of relationships

As there are 439120 observations, we will sample 5000 of them for visualisation and compute purposes.

```{r plots1, echo = FALSE}
train_sample_i <- sample(1:nrow(train), size = 5000)
train_sampled <- train[train_sample_i,]
pairs(train_sampled[,5:16])
#plot(train$TyreLife, train$Stint, main = "Tyre Life vs Stint", xlab = "TyreLife", ylab = "Stint")
```
Some cause for concern is the LapNumber and RaceProgression, as the latter is a function of the former, hence we will RaceProgression. All other variables are can be used. No clear sign of a need for transforming the variables but we can look into that later.

```{r cleaning, echo = FALSE}
train_sampled <- train_sampled[,-14]
```

## Modelling and predictions

Let's go for a simple model of predicting the whether we pit on the next lap using LapNumber, TyreLife and Cumulative_Degradation. We will then predict on another sample of 5000 observations for testing purposes and compute accuracy score for this simple baseline model.

```{r modeling}

library(dplyr)

logistic_model <- glm(PitNextLap ~ LapNumber + TyreLife,
                    data = train_sampled,
                    family = "binomial")
logistic_model

summary(logistic_model)
```

```{r precitions, echo = FALSE}
predict_reg <- predict(logistic_model,
                       train_sampled, type = "response")
for(i in 1:length(predict_reg)){
  pred <- predict_reg[i]
  if(pred >= 0.5){
    predict_reg[i] <- 1
  }
  else{
    predict_reg[i] <- 0
  }
}
predict_reg <- as.data.frame(as.factor(predict_reg))
```

```{r confusionMatrix, echo = FALSE}
library(ggplot2)
library(reshape2)

conf_matrix <- table(train_sampled$PitNextLap, predict_reg$`as.factor(predict_reg)`)

# Reshape the confusion matrix for ggplot2
conf_matrix_melted <- as.data.frame(conf_matrix)
colnames(conf_matrix_melted) <- c("Actual", "Predicted", "Count")

ggplot(conf_matrix_melted, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "black", size = 6) +  # Add text labels
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()
```

```{r accuracy, echo = FALSE}
accuracy_score <- sum(train_sampled$PitNextLap == predict_reg$`as.factor(predict_reg)`)/5000

paste("Accuracy Score of:", accuracy_score)
```
A good baseline prediction accuracy of 0.7944. We will now do a model with all of the variables included and remove all statistically insignificant variables, one at a time.
```{r modelling2, echo = FALSE}
logistic_model2 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))
summary(logistic_model2)
```

```{r prediction, include=FALSE}
logistic_model3 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))
summary(logistic_model3)
```

```{r prediction3, include=FALSE}
train_sampled <- train_sampled[,-c(10)]
logistic_model4 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))
summary(logistic_model4)
```

```{r prediction4, include=FALSE}
train_sampled <- train_sampled[,-c(11)]
logistic_model5 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))
summary(logistic_model5)
```

```{r prediction5, include=FALSE}
train_sampled <- train_sampled[,-c(10)]
logistic_model6 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))
summary(logistic_model6)
```

```{r prediction2, echo = FALSE}
predict_reg2 <- predict(logistic_model6,
                       train_sampled, type = "response")
for(i in 1:length(predict_reg2)){
  pred <- predict_reg2[i]
  if(pred > 0.5){
    predict_reg2[i] <- 1
  }
  else{
    predict_reg2[i] <- 0
  }
}
predict_reg2 <- as.data.frame(as.factor(predict_reg2))


conf_matrix <- table(train_sampled$PitNextLap, predict_reg2[[1]])

# Reshape the confusion matrix for ggplot2
conf_matrix_melted <- as.data.frame(conf_matrix)
colnames(conf_matrix_melted) <- c("Actual", "Predicted", "Count")

ggplot(conf_matrix_melted, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "black", size = 6) +  # Add text labels
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()

accuracy_score <- sum(train_sampled$PitNextLap == as.vector(predict_reg2[[1]]))/5000

paste("Accuracy Score of:", accuracy_score)
```

Now that we have narrowed down the variables we can take a closer look at the relationships of our variables and see if a transformation can improve our predictions.

```{r visualisations, echo = FALSE}
pairs(train_sampled[,-c(1:4)])
```
TyreLife appears to be quite spread so a we will log transform this and re run our model.

```{r transform, echo = FALSE}
train_sampled$TyreLife <- log(train_sampled$TyreLife)

logistic_model7 <- glm(PitNextLap ~ .,
                    data = train_sampled[,-c(1,2)],
                    family = binomial(link = "logit"))


predict_reg2 <- predict(logistic_model7,
                       train_sampled, type = "response")
for(i in 1:length(predict_reg2)){
  pred <- predict_reg2[i]
  if(pred > 0.5){
    predict_reg2[i] <- 1
  }
  else{
    predict_reg2[i] <- 0
  }
}
predict_reg2 <- as.data.frame(as.factor(predict_reg2))


conf_matrix <- table(train_sampled$PitNextLap, predict_reg2[[1]])

# Reshape the confusion matrix for ggplot2
conf_matrix_melted <- as.data.frame(conf_matrix)
colnames(conf_matrix_melted) <- c("Actual", "Predicted", "Count")

ggplot(conf_matrix_melted, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "black", size = 6) +  # Add text labels
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()

accuracy_score <- sum(train_sampled$PitNextLap == as.vector(predict_reg2[[1]]))/5000

paste("Accuracy Score of:", accuracy_score)
```



It looks like log transformation was a good idea. Improving our accuracy score by 1%. 


```{r submission, echo = FALSE}
test <- read.csv("test.csv")
test_cleaned <- test[ ,-c(15,12, 11,10)]
test_cleaned$Compound <- as.factor(test_cleaned$Compound)
test_cleaned$Race <- as.factor(test_cleaned$Race)
test_cleaned$Year <- as.factor(test_cleaned$Year)
test_cleaned$PitStop <- as.factor(test_cleaned$PitStop)
test_cleaned$TyreLife <- log(test_cleaned$TyreLife)

train_cleaned <- train[ ,-c(15,12, 11,10)]
train_cleaned$TyreLife <- log(train_cleaned$TyreLife)

logistic_model8 <- glm(PitNextLap ~ .,
                    data = train_cleaned[,-c(1,2)],
                    family = binomial(link = "logit"))

predict_reg2 <- predict(logistic_model8,
                       test_cleaned, type = "response")
submissiofileLogistic <- data.frame(id = test_cleaned$id, PitNextLap = predict_reg2)
write.csv(submissiofileLogistic, file = "F1SubmissionLogistic.csv", row.names = FALSE)
```

Our Logistic Regression Classifier resulted in a 89.27 AUC score on the Kaggle Test set, very good result for a baseline model.

## Random Forests

Now that we have exhausted the capabilities of logistic regression, we will move on to more complex methods, starting with Random Forests. We will start with a default number of trees 500 and mtry (the number of randomly chosen variables to choose from when deciding which variable to split from) chosen to be the square root of the number of variables, this being the most common practice and recommendation when working with random forests.


```{r RandomForest, echo = FALSE}
# Install if needed: 
#install.packages(c("randomForest", "caret", "pROC"))
library(randomForest)
library(pROC)
train <- read.csv("train.csv")
# Using the same prepared mtcars data from before
set.seed(123)  # for reproducibility

train_idx <- sample(1:nrow(train), size = round(nrow(train)*0.7))
test_idx <- seq(1,nrow(train))
test_idx <- test_idx[-train_idx]
trainlocal <- train[train_idx, ]
testlocal <- train[-train_idx,]
```

```{r RandomForests2, echo = FALSE}

# Fit Random Forest
rf_model <- randomForest(PitNextLap ~ ., 
                         data = trainlocal[,-c(1,2)], 
                         ntree = 500,          # number of trees
                         mtry = 4,             # variables tried at each split
                         importance = TRUE)

# Predictions on test set
rf_pred_prob <- predict(rf_model, testlocal, type = "prob")[, 2]  # probability of PitNextLap=1
rf_pred_class <- predict(rf_model, testlocal)                     # class prediction

# Accuracy

conf_matrix <- table(testlocal$PitNextLap, rf_pred_class)

# Reshape the confusion matrix for ggplot2
conf_matrix_melted <- as.data.frame(conf_matrix)
colnames(conf_matrix_melted) <- c("Actual", "Predicted", "Count")

ggplot(conf_matrix_melted, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "black", size = 6) +  # Add text labels
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()

accuracy_score <- sum(testlocal$PitNextLap == rf_pred_class)/nrow(testlocal)

paste("Accuracy Score of:", accuracy_score)
# ROC-AUC
roc_rf <- roc(testlocal$PitNextLap, rf_pred_prob)
print(auc(roc_rf))
plot(roc_rf, main = "Random Forest ROC Curve")

# Feature importance
importance(rf_model)
varImpPlot(rf_model)
```
The Random Forest proved to be a better predictor than logistic regression.

```{r RF_Submission, echo = FALSE}
test$Compound <- as.factor(test$Compound)
test$Race <- as.factor(test$Race)
test$Year <- as.factor(test$Year)
test$PitStop <- as.factor(test$PitStop)
rf_pred_class_Submission <- predict(rf_model, test[, -c(1,2)], type = "prob")[, 2] 
submissiofilerf <- data.frame(id = test$id, PitNextLap = rf_pred_class_Submission)
write.csv(submissiofilerf, file = "F1SubmissionRandomForest.csv", row.names = FALSE)
```
Time to tune the Random Forest. 
We have a huge class imbalance, roughly 80/20. This needs to be taken into consideration when modelling as we can get 80% accuracy if we just predict all observations to not pit on the next lap, but this leads to inaccurate and a false confidence in our predictions. Due to the imbalance we have a high type ii error, false positives and so our metric we need to improve is recall instead of accuracy.

```{r RFTuned, echo = FALSE}
# Fit Random Forest
rf_model <- randomForest(PitNextLap ~ ., 
                         data = trainlocal[,-c(1,2)], 
                         ntree = 800,          # number of trees
                         mtry = 4,
                         sampsize = c(0.7 * sum(trainlocal$PitNextLap == 0),   # sample this many from majority
                                      0.7 * sum(trainlocal$PitNextLap == 1)),  # sample almost all from minority             # variables tried at each split
                         strata = trainlocal$PitNextLap,
                         importance = TRUE)

class_counts <- table(trainlocal$PitNextLap)          # replace 'target' with your actual column
class_weights <- length(trainlocal$PitNextLap) / (2 * class_counts)

final_rf <- randomForest(PitNextLap ~ ., 
                         data = trainlocal[-c(1,2)],
                         ntree = 500,
                         mtry = 4,               # tune this
                         classwt = class_weights,   # <-- key line
                         importance = TRUE)
# Predictions on test set
rf_pred_prob <- predict(rf_model, testlocal, type = "prob")[, 2]  # probability of PitNextLap=1
rf_pred_class <- predict(rf_model, testlocal)                     # class prediction

print(final_rf)
# Accuracy

conf_matrix <- table(testlocal$PitNextLap, rf_pred_class)

# Reshape the confusion matrix for ggplot2
conf_matrix_melted <- as.data.frame(conf_matrix)
colnames(conf_matrix_melted) <- c("Actual", "Predicted", "Count")

ggplot(conf_matrix_melted, aes(x = Actual, y = Predicted, fill = Count)) +
  geom_tile() +
  geom_text(aes(label = Count), color = "black", size = 6) +  # Add text labels
  scale_fill_gradient(low = "white", high = "blue") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual", y = "Predicted") +
  theme_minimal()

accuracy_score <- sum(testlocal$PitNextLap == rf_pred_class)/nrow(testlocal)
true_positives <- testlocal$PitNextLap[which(testlocal$PitNextLap == 1)]
recall <- sum(true_positives == rf_pred_class[which(testlocal$PitNextLap == 1)])/length(true_positives)

paste("Accuracy Score of:", accuracy_score)
paste("Recall Score of:", recall)
# ROC-AUC
roc_rf <- roc(testlocal$PitNextLap, rf_pred_prob)
print(auc(roc_rf))
plot(roc_rf, main = "Random Forest ROC Curve")

# Feature importance
importance(rf_model)
varImpPlot(rf_model)
```

```{r RF2, echo = FALSE}
n_trees = c(500,750)
mtry_list <- c(3,4,5)
train$Driver <- as.factor(train$Driver)
train_sampled <- train[sample(20000), ]
accuracy_matrix <- matrix(0, nrow = 5, ncol = length(n_trees)*length(mtry_list))
j <- 1
l <- 1

class_counts <- table(train$PitNextLap)          # replace 'target' with your actual column
class_weights <- length(train$PitNextLap) / (2 * class_counts)

for(n_tree in n_trees){
  for(mt in mtry_list){
    # 1. Setup
    set.seed(123) # For reproducibility
    k <- 5
    data <- train_sampled[sample(nrow(train_sampled)), ] # Shuffle data

    # 2. Create Fold Indices
    folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
    errors <- numeric(k)
    
    # 3. Perform Cross-Validation
    for(i in 1:k) {
        # Split into train and test
        test_indices <- which(folds == i, arr.ind = TRUE)
        test_data <- data[test_indices, ]
        train_data <- data[-test_indices, ]
        
        rf_model <- randomForest(PitNextLap ~ ., 
                         data = train_data[,-c(1,2)], 
                         ntree = n_tree,          # number of trees
                         mtry = mt, 
                         classwt = class_weights,
                         # variables tried at each split
                         importance = TRUE)
        rf_pred_class <- predict(rf_model, test_data) 

        accuracy_matrix[i,l] <- sum(test_data$PitNextLap == rf_pred_class)/nrow(test_data)
        accuracy_score <- sum(test_data$PitNextLap == rf_pred_class)/nrow(test_data)
        true_positives <- test_data$PitNextLap[which(test_data$PitNextLap == 1)]
        recall <- sum(true_positives == rf_pred_class[which(test_data$PitNextLap == 1)])/length(true_positives)

        cat("Accuracy Score of:", accuracy_score, "ntree:",n_tree, "mtry:", mt, "\n")
        cat("Recall Score of:", recall, "\n")

    }
    names(accuracy_matrix)[l] <- paste0(n_tree,"plus",mt)
    l <- l + 1


  }
}

accuracy_matrix
```
The class imbalance is still an issue as our recall is still hovering around the 63% range, we should ideally bring this up above 70% so we will test different thresholds for prediction to see which improves recall the most. We will use a RF with 500 trees and mtry of 4 as this turns out to be the best as initially chosen.

```{r recallimprovement, echo = FALSE}
# 1. Setup
set.seed(123) # For reproducibility
k <- 5
data <- train_sampled[sample(nrow(train_sampled)), ] # Shuffle data

# 2. Create Fold Indices
folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
errors <- numeric(k)
prediction_mat <- matrix(0, nrow = k, ncol = length(thresholds)*2)
# 3. Perform Cross-Validation
for(i in 1:k) {
    # Split into train and test
    test_indices <- which(folds == i, arr.ind = TRUE)
    test_data <- data[test_indices, ]
    train_data <- data[-test_indices, ]
    
    rf_model <- randomForest(PitNextLap ~ ., 
                     data = train_data[,-c(1,2)], 
                     ntree = 500,          # number of trees
                     mtry = 4, 
                     classwt = class_weights,
                     # variables tried at each split
                     importance = TRUE)
    pred_prob <- predict(rf_model, test_data, type = "prob")[, "1"] 
    # Try different thresholds
    thresholds <- seq(0.25, 0.55, by = 0.05)
    thcol <- 1
    for(th in thresholds){
      pred_class <- ifelse(pred_prob > th, 1, 0)
      accuracy_score <- sum(test_data$PitNextLap == pred_class)/nrow(test_data)
      true_positives <- test_data$PitNextLap[which(test_data$PitNextLap == 1)]
      recall <- sum(true_positives == pred_class[which(test_data$PitNextLap == 1)])/length(true_positives)
      prediction_mat[i, thcol] <- accuracy_score
      prediction_mat[i, thcol+1] <- recall
      cat("Accuracy Score of:", accuracy_score, "threshold:",th, "\n")
      cat("Recall Score of:", recall, "\n")
      thcol <- thcol + 2
    }

}

```

```{r plotthreshold, echo = FALSE}
predictionmatmean <- colMeans(prediction_mat)
accuracymean <- predictionmatmean[seq(1,14, by = 2)]
recallmean <- predictionmatmean[seq(2,14, by = 2)]

plot(accuracymean, recallmean, main = "accuracy vs recall for differing prediciton thresholds")
predictionmatmean
```
As we can see, with a threshold of 0.25 our recall is very high (~0.85) and a very similar accuracy (~0.865) but we are missing out on 1-2% overall accuracy, but as seen in the test set on kaggle, it appears that the test set has more positives than in our training set, hence a higher recall may be a good idea. Lets run on the full data with this threshold and see how it plays out.

```{r recallimprovedsubmissionRF, echo = FALSE}
test$Compound <- as.factor(test$Compound)
test$Race <- as.factor(test$Race)
test$Year <- as.factor(test$Year)
test$PitStop <- as.factor(test$PitStop)
test$Driver <- as.factor(test$Driver)
library(randomForest)
rf_model <- randomForest(PitNextLap ~ ., 
                 data = train[,-c(1,2)], 
                 ntree = 500,          # number of trees
                 mtry = 4, 
                 classwt = class_weights,
                 # variables tried at each split
                 importance = TRUE)
print(rf_model)
plot(rf_model)
importance(rf_model)
varImpPlot(rf_model)
pred_prob <- predict(rf_model, test, type = "prob")[, "1"] 
submissiofileRF <- data.frame(id = test$id, PitNextLap = pred_prob)
write.csv(submissiofileRF, file = "F1SubmissionRF_RecallImproved.csv", row.names = FALSE)
```
As we can see from the variable importance plot, the race track was the most important in terms of increasing the accuracy, closely followed by Stint, Tyrelife and Year, as when we drop these variable separately, we see the largest mean drop in accuracy. That is interesting as you would not imagine that the race track itself would be the most important feature, intuitively the Stint would be the most important feature but different track, wear the tyres in different ways, some track will wear out the tyres more and or require different tyre change tactics that are not solely dependent on tyre wear, but more about a tactical pit to get out of traffic so that the driver has free air in front of him to set fast laps.


Perhaps we can manufacture a variable called rollavgdelta. This variable would be a rolling average lap time delta of the last 5 laps. Why I believe this could be useful is the fact that a driver may decide to pit or the team will make the driver pit if they notice a drop off in lap times consistently and so if the rolling average delta is positive, this would indicate there pace is dropping and is likely due to tyre wear and an indication of a need to pit. 

There is an issue with doing this as the training data is sampled from the full data, hence we do not have consecutive laps all of the time. Hence we only calculate the rolling average where the laps are consecutive for at least 3 laps.


```{r eval=FALSE, include=FALSE, rolling_avg_lap_delta, echo=FALSE}
library(data.table)
library(dplyr)

create_consecutive_rolling_features <- function(data) {
  setDT(data)
  
  # Ensure correct ordering
  data <- data[order(Driver, Race, Year, LapNumber)]
  
  # Create a group identifier for consecutive lap streaks
  data[, LapDiff := LapNumber - shift(LapNumber, 1, fill = NA),
       by = .(Driver, Race, Year)]
  
  data[, StreakID := cumsum(LapDiff != 1 | is.na(LapDiff)),
       by = .(Driver, Race, Year)]
  
  # Calculate LapTimeDelta only within streak
  data[, LapTimeDelta := LapTime_Delta,
       by = .(Driver, Race, Year, StreakID)]
  
  # Rolling features — only when streak has enough laps
  data[, `:=`(
    RollingAvgDelta_5 = ifelse(.N >= 3,
                               frollmean(LapTimeDelta, n = 5, align = "right", na.rm = TRUE),
                               0),
    RollingAvgDelta_3 = ifelse(.N >= 3,
                               frollmean(LapTimeDelta, n = 3, align = "right", na.rm = TRUE),
                               0),
    RollingAvgDelta_10 = ifelse(.N >= 5,
                                frollmean(LapTimeDelta, n = 10, align = "right", na.rm = TRUE),
                                0),
    RollingStdDelta_5 = ifelse(.N >= 4,
                               frollapply(LapTimeDelta, n = 5, FUN = sd, na.rm = TRUE),
                               0)
  ), by = .(Driver, Race, Year, StreakID)]
  
  # Additional useful features
  data[, `:=`(
    TireDegradationTrend = RollingAvgDelta_5 * TyreLife,     # interaction
    IsDegradingFast = as.integer(RollingAvgDelta_5 > 0.3),   # tunable threshold
    RecentPaceDrop = RollingAvgDelta_3 > 0.15
  )]
  
  # Clean up temporary columns
  data[, c("LapDiff", "StreakID") := NULL]
  
  return(as_tibble(data))   # convert back to tibble if you prefer
}


# Apply it
trainrolling <- create_consecutive_rolling_features(train)
summary(trainrolling)
testrolling  <- create_consecutive_rolling_features(test)

trainrolling$RollingStdDelta_5[is.na(trainrolling$RollingStdDelta_5)] <- 0
testrolling$RollingStdDelta_5[is.na(testrolling$RollingStdDelta_5)] <- 0
trainrolling$RollingAvgDelta_10[is.na(trainrolling$RollingAvgDelta_10)] <- 0
testrolling$RollingAvgDelta_10[is.na(testrolling$RollingAvgDelta_10)] <- 0
trainrolling$RollingAvgDelta_5[is.na(trainrolling$RollingAvgDelta_5)] <- 0
testrolling$RollingAvgDelta_5[is.na(testrolling$RollingAvgDelta_5)] <- 0
```

This code does not work as we do not have consecutive laps to work with in a meaningful way.

## XGBoost

```{r XGBoost, echo = FALSE}
# =============================================
# XGBoost Starter Code for Binary Classification
# With class imbalance handling (80/20) + mixed predictors
# =============================================
#install.packages("xgboost", 
#                type = "binary", 
#                 repos = "https://cran.rstudio.com/")
#install.packages("caret", 
#                 type = "binary", 
#                 repos = "https://cran.rstudio.com/")
library(xgboost)
library(caret)      # for dummyVars + train/test split
library(pROC)       # for AUC
library(dplyr)
library(data.table)

# Assume you already have:
# train and test data frames
# target column called 'target' (0/1)

# =============================================
# CLEANED XGBoost Training Block
# =============================================

# Make sure target is 0/1 numeric
trainlocal$PitNextLap <- as.numeric(trainlocal$PitNextLap)# Important: 0 and 1
testlocal$PitNextLap  <- as.numeric(testlocal$PitNextLap) 

# Identify categorical variables
cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]

# ====================== FEATURE PREPARATION ======================
if(length(cat_vars) > 0){
  
  # Train dummies
  cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
  dummies <- dummyVars(~ ., data = cat_data_train)
  dummy_train <- predict(dummies, cat_data_train)
  
  # Numeric columns
  num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
  X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
  
  # Test dummies (using same dummy structure)
  cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
  dummy_test <- predict(dummies, cat_data_test)
  X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
  
} else {
  X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
  X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
}
# ====================== FIX FEATURE ALIGNMENT ======================

# Make sure test has exactly the same columns as train (in same order)
common_cols <- intersect(names(X_train), names(X_test))

# Add missing columns to test (filled with 0)
missing_in_test <- setdiff(names(X_train), names(X_test))
for (col in missing_in_test) {
  X_test[[col]] <- 0
}

# Add missing columns to train (just in case)
missing_in_train <- setdiff(names(X_test), names(X_train))
for (col in missing_in_train) {
  X_train[[col]] <- 0
}

# Re-order test columns to match train exactly
X_test <- X_test[, names(X_train), drop = FALSE]

# Now create DMatrix again
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)

cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")

# ====================== CLASS IMBALANCE ======================
pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
cat("scale_pos_weight:", round(pos_weight, 2), "\n")

# ====================== CHECK CLASS DISTRIBUTION ======================
cat("Train class distribution:\n")
print(table(trainlocal$PitNextLap))
cat("Test class distribution:\n")
print(table(testlocal$PitNextLap))

# ====================== MODEL TRAINING ======================
params <- list(
  objective         = "binary:logistic",
  eval_metric       = "logloss",          # safe choice
  eta               = 0.05,
  max_depth         = 6,
  subsample         = 0.8,
  colsample_bytree  = 0.8,
  scale_pos_weight  = pos_weight,
  gamma             = 1,
  min_child_weight  = 1,
  nthread           = parallel::detectCores() - 1
)

set.seed(123)

xgb_model <- xgb.train(
  params            = params,
  data              = dtrain,
  nrounds           = 500,
  watchlist         = list(train = dtrain, test = dtest),
  early_stopping_rounds = 50,
  verbose           = 10
)

# ====================== PREDICTIONS ======================
pred_prob <- predict(xgb_model, dtest)
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

# Evaluation
library(caret)
cm <- confusionMatrix(factor(pred_class, levels = c(0,1)),
                      factor(testlocal$PitNextLap, levels = c(0,1)),
                      positive = "1")
print(cm)
```

```{r XGBoostSubmission, echo = FALSE}

# =============================================
# XGBoost Starter Code for Binary Classification
# With class imbalance handling (80/20) + mixed predictors
# =============================================


# Assume you already have:
# train and test data frames
# target column called 'target' (0/1)

# =============================================
# CLEANED XGBoost Training Block
# =============================================

train_idx <- sample(1:nrow(train), size = round(nrow(train)*0.7))
trainlocal <- train[train_idx,-c(1,2)]
testlocal <- train[-train_idx,-c(1,2)]

# Make sure target is 0/1 numeric
trainlocal$PitNextLap <- as.numeric(trainlocal$PitNextLap) - 1# Important: 0 and 1
testlocal$PitNextLap <- as.numeric(testlocal$PitNextLap) - 1

# Identify categorical variables
cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]

# ====================== FEATURE PREPARATION ======================
if(length(cat_vars) > 0){
  
  # Train dummies
  cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
  dummies <- dummyVars(~ ., data = cat_data_train)
  dummy_train <- predict(dummies, cat_data_train)
  
  # Numeric columns
  num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
  X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
  
  # Test dummies (using same dummy structure)
  cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
  dummy_test <- predict(dummies, cat_data_test)
  X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
  
} else {
  X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
  X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
}
# ====================== FIX FEATURE ALIGNMENT ======================

# Make sure test has exactly the same columns as train (in same order)
common_cols <- intersect(names(X_train), names(X_test))

# Add missing columns to test (filled with 0)
missing_in_test <- setdiff(names(X_train), names(X_test))
for (col in missing_in_test) {
  X_test[[col]] <- 0
}

# Add missing columns to train (just in case)
missing_in_train <- setdiff(names(X_test), names(X_train))
for (col in missing_in_train) {
  X_train[[col]] <- 0
}

# Re-order test columns to match train exactly
X_test <- X_test[, names(X_train), drop = FALSE]

# Now create DMatrix again
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)

cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")

# ====================== CLASS IMBALANCE ======================
pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
cat("scale_pos_weight:", round(pos_weight, 2), "\n")

# ====================== CHECK CLASS DISTRIBUTION ======================
cat("Train class distribution:\n")
print(table(trainlocal$PitNextLap))
cat("Test class distribution:\n")
print(table(testlocal$PitNextLap))

# ====================== MODEL TRAINING ======================
params <- list(
  objective         = "binary:logistic",
  eval_metric       = "logloss",          # safe choice
  eta               = 0.05,
  max_depth         = 6,
  subsample         = 0.8,
  colsample_bytree  = 0.8,
  scale_pos_weight  = pos_weight,
  gamma             = 1,
  min_child_weight  = 1,
  nthread           = parallel::detectCores() - 1
)

set.seed(123)

xgb_model <- xgb.train(
  params            = params,
  data              = dtrain,
  nrounds           = 500,
  watchlist         = list(train = dtrain, test = dtest),
  early_stopping_rounds = 50,
  verbose           = 10
)

# ====================== PREDICTIONS ======================
# 
  # Test dummies (using same dummy structure)
cat_data_test <- as.data.frame(test[, cat_vars, drop = FALSE])
dummy_test <- predict(dummies, cat_data_test)
X_test <- cbind(as.data.frame(test[, num_cols, drop = FALSE]), dummy_test)
#test$PitFinalLap <- as.numeric(test$PitFinalLap) - 1
dtestfinal  <- xgb.DMatrix(data = as.matrix(X_test))
pred_prob <- predict(xgb_model, dtestfinal)

# Evaluation
library(caret)
cm <- confusionMatrix(factor(pred_class, levels = c(0,1)),
                      factor(testlocal$PitNextLap, levels = c(0,1)),
                      positive = "1")
print(cm)


submissionfileXG <- data.frame(id = test$id, PitNextLap = pred_prob)
write.csv(submissionfileXG, file = "F1SubmissionXG.csv", row.names = FALSE)
```

First run on XGBoost resulted in a 0.87941 accuracy score on the kaggle test set. Not bad but lets improve our prediction and do some tuning.

```{r XGBoostTuning, echo = FALSE}

traindata <- train[train_idx,-c(1,2)]
testdata<- train[-train_idx,-c(1,2)]
traindata$PitNextLap <- as.numeric(traindata$PitNextLap) - 1# Important: 0 and 1
testdata$PitNextLap  <- as.numeric(testdata$PitNextLap) - 1

set.seed(123) # For reproducibility
k <- 5
data <- traindata

# 2. Create Fold Indices
folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
errors <- numeric(k)
thresholds <- seq(0.25, 0.55, by = 0.05)
prediction_mat <- matrix(0, nrow = k, ncol = length(thresholds)*2)
# 3. Perform Cross-Validation
for(i in 1:k) {
    # Split into train and test
    test_indices <- which(folds == i, arr.ind = TRUE)
    testlocal <- data[test_indices, ]
    trainlocal <- data[-test_indices, ]
    
    # Identify categorical variables
    cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]
    
    # ====================== FEATURE PREPARATION ======================
    if(length(cat_vars) > 0){
      
      # Train dummies
      cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
      dummies <- dummyVars(~ ., data = cat_data_train)
      dummy_train <- predict(dummies, cat_data_train)
      
      # Numeric columns
      num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
      X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
      
      # Test dummies (using same dummy structure)
      cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
      dummy_test <- predict(dummies, cat_data_test)
      X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
      
    } else {
      X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
      X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
    }
    # ====================== FIX FEATURE ALIGNMENT ======================
    
    # Make sure test has exactly the same columns as train (in same order)
    common_cols <- intersect(names(X_train), names(X_test))
    
    # Add missing columns to test (filled with 0)
    missing_in_test <- setdiff(names(X_train), names(X_test))
    for (col in missing_in_test) {
      X_test[[col]] <- 0
    }
    
    # Add missing columns to train (just in case)
    missing_in_train <- setdiff(names(X_test), names(X_train))
    for (col in missing_in_train) {
      X_train[[col]] <- 0
    }
    
    # Re-order test columns to match train exactly
    X_test <- X_test[, names(X_train), drop = FALSE]
    
    # Now create DMatrix again
    dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
    dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)
    
    cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")
    
    # ====================== CLASS IMBALANCE ======================
    pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
    cat("scale_pos_weight:", round(pos_weight, 2), "\n")
    
    # ====================== CHECK CLASS DISTRIBUTION ======================
    cat("Train class distribution:\n")
    print(table(trainlocal$PitNextLap))
    cat("Test class distribution:\n")
    print(table(testlocal$PitNextLap))
    
    # ====================== MODEL TRAINING ======================
    params <- list(
      objective         = "binary:logistic",
      eval_metric       = "logloss",          # safe choice
      eta               = 0.05,
      max_depth         = 6,
      subsample         = 0.8,
      colsample_bytree  = 0.8,
      scale_pos_weight  = pos_weight,
      gamma             = 1,
      min_child_weight  = 1,
      nthread           = parallel::detectCores() - 1
    )
    
    set.seed(123)
    
    xgb_model <- xgb.train(
      params            = params,
      data              = dtrain,
      nrounds           = 500,
      watchlist         = list(train = dtrain, test = dtest),
      early_stopping_rounds = 50,
      verbose           = 10
    )
    
    # ====================== PREDICTIONS ======================
    pred_prob <- predict(xgb_model, dtest)
    # Evaluation

    # Try different thresholds
    thcol <- 1
    for(th in thresholds){
      pred_class <- ifelse(pred_prob > th, 1, 0)
      accuracy_score <- sum(testlocal$PitNextLap == pred_class)/nrow(testlocal)
      true_positives <- testlocal$PitNextLap[which(testlocal$PitNextLap == 1)]
      recall <- sum(true_positives == pred_class[which(testlocal$PitNextLap == 1)])/length(true_positives)
      prediction_mat[i, thcol] <- accuracy_score
      prediction_mat[i, thcol+1] <- recall
      cat("Accuracy Score of:", accuracy_score, "threshold:",th, "\n")
      cat("Recall Score of:", recall, "\n")
      thcol <- thcol + 2
    }

}

predictionmeans <- colMeans(prediction_mat)
accuracymeans <- predictionmeans[seq(1,14,by=2)]
recallmeans <- predictionmeans[seq(2,14,by=2)]

plot(accuracymeans, recallmeans, main = "Accuracy vs Recall means for XGBoost with 5-fold CV")
```
Analysing the plot of recall means versus accuracy means, the optimal threshold appears to be around the 0.35 to 0.4 region, at which point we have a high recall of 91.88% and an accuracy score of 85.86%.

Due to the minor change in recall and accuracy from the original threshold of 0.5, I will hold off on submitting another attempt until I have tuned with other parameters but from here on in the threshold we will be testing with 0.35 to 0.45 thresholds.

```{r XGBoostFurtherTuning, echo = FALSE}

traindata <- train[train_idx,-c(1,2)]
testdata<- train[-train_idx,-c(1,2)]
traindata$PitNextLap <- as.numeric(traindata$PitNextLap) - 1# Important: 0 and 1
testdata$PitNextLap  <- as.numeric(testdata$PitNextLap) - 1

set.seed(123) # For reproducibility
k <- 5
data <- traindata

# 2. Create Fold Indices
folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
errors <- numeric(k)
thresholds <- seq(0.25, 0.55, by = 0.05)
prediction_mat <- matrix(0, nrow = k, ncol = length(thresholds)*2)
# 3. Perform Cross-Validation
for(i in 1:k) {
    # Split into train and test
    test_indices <- which(folds == i, arr.ind = TRUE)
    testlocal <- data[test_indices, ]
    trainlocal <- data[-test_indices, ]
    
    # Identify categorical variables
    cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]
    
    # ====================== FEATURE PREPARATION ======================
    if(length(cat_vars) > 0){
      
      # Train dummies
      cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
      dummies <- dummyVars(~ ., data = cat_data_train)
      dummy_train <- predict(dummies, cat_data_train)
      
      # Numeric columns
      num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
      X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
      
      # Test dummies (using same dummy structure)
      cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
      dummy_test <- predict(dummies, cat_data_test)
      X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
      
    } else {
      X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
      X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
    }
    # ====================== FIX FEATURE ALIGNMENT ======================
    
    # Make sure test has exactly the same columns as train (in same order)
    common_cols <- intersect(names(X_train), names(X_test))
    
    # Add missing columns to test (filled with 0)
    missing_in_test <- setdiff(names(X_train), names(X_test))
    for (col in missing_in_test) {
      X_test[[col]] <- 0
    }
    
    # Add missing columns to train (just in case)
    missing_in_train <- setdiff(names(X_test), names(X_train))
    for (col in missing_in_train) {
      X_train[[col]] <- 0
    }
    
    # Re-order test columns to match train exactly
    X_test <- X_test[, names(X_train), drop = FALSE]
    
    # Now create DMatrix again
    dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
    dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)
    
    cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")
    
    # ====================== CLASS IMBALANCE ======================
    pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
    cat("scale_pos_weight:", round(pos_weight, 2), "\n")
    
    # ====================== CHECK CLASS DISTRIBUTION ======================
    cat("Train class distribution:\n")
    print(table(trainlocal$PitNextLap))
    cat("Test class distribution:\n")
    print(table(testlocal$PitNextLap))
    
    # ====================== MODEL TRAINING ======================
    params <- list(
      objective         = "binary:logistic",
      eval_metric       = "auc",          # changed from logloss
      eta               = 0.03,           # decreased from 0.05 for a slower learning rate
      max_depth         = 7,              # increased from 6
      subsample         = 0.85,           # increased from 0.8
      colsample_bytree  = 0.75,           # increased from 0.8
      scale_pos_weight  = pos_weight*1.2, # factor of 1.2 increase
      gamma             = 0.5,            #  
      min_child_weight  = 3,
      max_delta_step    = 1,
      nthread           = parallel::detectCores() - 1
    )
    
    set.seed(123)
    
    xgb_model <- xgb.train(
      params            = params,
      data              = dtrain,
      nrounds           = 500,
      watchlist         = list(train = dtrain, test = dtest),
      early_stopping_rounds = 80,
      verbose           = 20
    )
    
    # ====================== PREDICTIONS ======================
    pred_prob <- predict(xgb_model, dtest)
    # Evaluation

    # Try different thresholds
    thcol <- 1
    for(th in thresholds){
      pred_class <- ifelse(pred_prob > th, 1, 0)
      accuracy_score <- sum(testlocal$PitNextLap == pred_class)/nrow(testlocal)
      true_positives <- testlocal$PitNextLap[which(testlocal$PitNextLap == 1)]
      recall <- sum(true_positives == pred_class[which(testlocal$PitNextLap == 1)])/length(true_positives)
      prediction_mat[i, thcol] <- accuracy_score
      prediction_mat[i, thcol+1] <- recall
      cat("Accuracy Score of:", accuracy_score, "threshold:",th, "\n")
      cat("Recall Score of:", recall, "\n")
      thcol <- thcol + 2
    }

}

predictionmeans <- colMeans(prediction_mat)
accuracymeans <- predictionmeans[seq(1,14,by=2)]
recallmeans <- predictionmeans[seq(2,14,by=2)]

plot(accuracymeans, recallmeans, main = "Accuracy vs Recall means for XGBoost with 5-fold CV")
```

```{r XGBoostTunedSubmission, echo = FALSE}
library(xgboost)
library(caret)      # for dummyVars + train/test split
library(pROC)       # for AUC
library(dplyr)
library(data.table)
train_idx <- sample(1:nrow(train), size = round(nrow(train)*0.7))
trainlocal <- train[train_idx,-c(1,2)]
testlocal <- train[-train_idx,-c(1,2)]

# Make sure target is 0/1 numeric
trainlocal$PitNextLap <- as.numeric(trainlocal$PitNextLap) - 1# Important: 0 and 1
testlocal$PitNextLap <- as.numeric(testlocal$PitNextLap) - 1

# Identify categorical variables
cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]

# ====================== FEATURE PREPARATION ======================
if(length(cat_vars) > 0){
  
  # Train dummies
  cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
  dummies <- dummyVars(~ ., data = cat_data_train)
  dummy_train <- predict(dummies, cat_data_train)
  
  # Numeric columns
  num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
  X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
  
  # Test dummies (using same dummy structure)
  cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
  dummy_test <- predict(dummies, cat_data_test)
  X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
  
} else {
  X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
  X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
}
# ====================== FIX FEATURE ALIGNMENT ======================

# Make sure test has exactly the same columns as train (in same order)
common_cols <- intersect(names(X_train), names(X_test))

# Add missing columns to test (filled with 0)
missing_in_test <- setdiff(names(X_train), names(X_test))
for (col in missing_in_test) {
  X_test[[col]] <- 0
}

# Add missing columns to train (just in case)
missing_in_train <- setdiff(names(X_test), names(X_train))
for (col in missing_in_train) {
  X_train[[col]] <- 0
}

# Re-order test columns to match train exactly
X_test <- X_test[, names(X_train), drop = FALSE]

# Now create DMatrix again
dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)

cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")

# ====================== CLASS IMBALANCE ======================
pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
cat("scale_pos_weight:", round(pos_weight, 2), "\n")

# ====================== CHECK CLASS DISTRIBUTION ======================
cat("Train class distribution:\n")
print(table(trainlocal$PitNextLap))
cat("Test class distribution:\n")
print(table(testlocal$PitNextLap))

# ====================== MODEL TRAINING ======================
params <- list(
  objective         = "binary:logistic",
  eval_metric       = "auc",          # changed from logloss
  eta               = 0.03,           # decreased from 0.05 for a slower learning rate
  max_depth         = 7,              # increased from 6
  subsample         = 0.85,           # increased from 0.8
  colsample_bytree  = 0.75,           # increased from 0.8
  scale_pos_weight  = pos_weight*1.2, # factor of 1.2 increase
  gamma             = 0.5,            #  
  min_child_weight  = 3,
  max_delta_step    = 1,
  nthread           = parallel::detectCores() - 1
)

set.seed(123)

xgb_model <- xgb.train(
  params            = params,
  data              = dtrain,
  nrounds           = 500,
  watchlist         = list(train = dtrain, test = dtest),
  early_stopping_rounds = 80,
  verbose           = 20
)
    
# ====================== PREDICTIONS ======================
# 
  # Test dummies (using same dummy structure)

cat_data_test <- as.data.frame(test[, cat_vars, drop = FALSE])
dummy_test <- predict(dummies, cat_data_test)
X_test <- cbind(as.data.frame(test[, num_cols, drop = FALSE]), dummy_test)
X_test <- as.matrix(X_test[, names(X_train), drop = FALSE])
#test$PitFinalLap <- as.numeric(test$PitFinalLap) - 1
dtestfinal  <- xgb.DMatrix(data = X_test)
pred_prob_xg <- predict(xgb_model, dtestfinal)
# Evaluation

submissionfileXG <- data.frame(id = test$id, PitNextLap = pred_prob_xg)
write.csv(submissionfileXG, file = "F1SubmissionXGTuned.csv", row.names = FALSE)
```

```{r NeuralNetwork, echo = FALSE}
# Create a new conda environment (recommended)
library(reticulate)
library(keras3)
library(dplyr)

# 1. Activate the existing environment
use_condaenv("r-reticulate", required = TRUE)

# 2. Install TensorFlow + Keras into this environment
install_keras(
  method = "conda",
  envname = "r-reticulate",     # use the existing one
  tensorflow = "2.17"           # stable version
)

library(keras3)
library(reticulate)
use_condaenv("r-reticulate", required = TRUE)

# Import tensorflow explicitly
tf <- import("tensorflow")

# ====================== DATA PREPARATION ======================
X_train_scaled <- as.matrix(X_train)
X_test_scaled  <- as.matrix(X_test)

y_train <- as.matrix(as.factor(as.numeric(trainlocal$PitNextLap)-1))
y_test  <- as.matrix(as.factor(as.numeric(testlocal$PitNextLap)-1))

pos_weight <- sum(y_train == 0) / sum(y_train == 1)

cat("Positive weight:", round(pos_weight, 2), "\n")

# ====================== BUILD MODEL ======================
model <- keras_model_sequential()

model$add(layer_dense(units = 256, activation = "relu", 
                      input_shape = ncol(X_train_scaled)))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.4))

model$add(layer_dense(units = 128, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.35))

model$add(layer_dense(units = 64, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.3))

model$add(layer_dense(units = 1, activation = "sigmoid"))

# ====================== COMPILE ======================
model$compile(
  optimizer = optimizer_adam(learning_rate = 0.001),
  loss = "binary_crossentropy",
  metrics = c("accuracy", metric_auc(name = "auc"))
)

summary(model)
class_weights <- reticulate::dict("0" = 1.0, "1" = pos_weight * 1.3)
# ====================== TRAIN ======================
history <- model$fit(
  x = tf$convert_to_tensor(X_train_scaled, dtype = "float32"),
  y = tf$convert_to_tensor(y_train, dtype = "float32"),
  epochs = as.integer(60),
  batch_size = as.integer(1024),
  validation_data = list(
    tf$convert_to_tensor(X_test_scaled, dtype = "float32"),
    tf$convert_to_tensor(y_test, dtype = "float32")
  ),
  class_weight = class_weights,
  callbacks = list(
    callback_early_stopping(
      monitor = "val_auc",
      patience = 15,
      restore_best_weights = TRUE,
      mode = "max"
    )
  ),
  verbose = 1
)

# ====================== PREDICTIONS ======================
pred_prob <- model$predict(
  tf$convert_to_tensor(X_test_scaled, dtype = "float32")
) %>% as.vector()

# Threshold tuning
thresholds <- seq(0.25, 0.60, by = 0.01)

results <- data.frame()

for(th in thresholds){
  pred_class <- ifelse(pred_prob > th, 1, 0)
  
  cm <- confusionMatrix(
    factor(pred_class, levels = c(0,1)),
    factor(testlocal$PitNextLap, levels = c(0,1)), 
    positive = "1"
  )
  
  results <- rbind(results, data.frame(
    threshold = th,
    recall    = cm$byClass["Recall"],
    precision = cm$byClass["Precision"],
    f1        = cm$byClass["F1"]
  ))
}

print(results[order(results$f1, decreasing = TRUE), ][1:8, ])

best_th <- results$threshold[which.max(results$f1)]
cat("\nBest threshold (by F1):", best_th, "\n")

```

Amazing results here with perfect precision and near perfect recall and f1 scores. I will now run this on the full data set and submit to kaggle.

```{r NeuralNetSubmission, echo = FALSE}
library(keras3)
library(reticulate)

use_condaenv("r-reticulate", required = TRUE)

install_keras(
  method = "conda",
  envname = "r-reticulate",     # use the existing one
  tensorflow = "2.17"           # stable version
)

library(keras3)
library(reticulate)

use_condaenv("r-reticulate", required = TRUE)

# Import tensorflow explicitly
tf <- import("tensorflow")

train_idx <- sample(1:nrow(train), size = round(nrow(train)*0.7))
trainlocal <- train[train_idx,-c(1,2)]
testlocal <- train[-train_idx,-c(1,2)]

# Make sure target is 0/1 numeric
trainlocal$PitNextLap <- as.numeric(trainlocal$PitNextLap)-1 # Important: 0 and 1
testlocal$PitNextLap <- as.numeric(testlocal$PitNextLap) -1

# Identify categorical variables
cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]

# ====================== FEATURE PREPARATION ======================
if(length(cat_vars) > 0){
  
  # Train dummies
  cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
  dummies <- dummyVars(~ ., data = cat_data_train)
  dummy_train <- predict(dummies, cat_data_train)
  
  # Numeric columns
  num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
  X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
  
  # Test dummies (using same dummy structure)
  cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
  dummy_test <- predict(dummies, cat_data_test)
  X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
  
} else {
  X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
  X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
}
# ====================== FIX FEATURE ALIGNMENT ======================

# Make sure test has exactly the same columns as train (in same order)
common_cols <- intersect(names(X_train), names(X_test))

# Add missing columns to test (filled with 0)
missing_in_test <- setdiff(names(X_train), names(X_test))
for (col in missing_in_test) {
  X_test[[col]] <- 0
}

# Add missing columns to train (just in case)
missing_in_train <- setdiff(names(X_test), names(X_train))
for (col in missing_in_train) {
  X_train[[col]] <- 0
}

# Re-order test columns to match train exactly
X_test <- X_test[, names(X_train), drop = FALSE]

tf <- import("tensorflow")

# ====================== DATA PREPARATION ======================
X_train_scaled <- as.matrix(X_train)
X_test_scaled  <- as.matrix(X_test)

y_train <- as.matrix(as.factor(as.numeric(trainlocal$PitNextLap)))
y_test  <- as.matrix(as.factor(as.numeric(testlocal$PitNextLap)))

pos_weight <- sum(y_train == 0) / sum(y_train == 1)

cat("Positive weight:", round(pos_weight, 2), "\n")

# ====================== BUILD MODEL ======================
model <- keras_model_sequential()

model$add(layer_dense(units = 256, activation = "relu", 
                      input_shape = ncol(X_train_scaled)))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.4))

model$add(layer_dense(units = 128, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.35))

model$add(layer_dense(units = 64, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.3))

model$add(layer_dense(units = 1, activation = "sigmoid"))

# ====================== COMPILE ======================
model$compile(
  optimizer = optimizer_adam(learning_rate = 0.001),
  loss = "binary_crossentropy",
  metrics = c("accuracy", metric_auc(name = "auc"))
)

summary(model)
class_weights <- reticulate::dict("0" = 1.0, "1" = pos_weight * 1.3)
# ====================== TRAIN ======================
history <- model$fit(
  x = tf$convert_to_tensor(X_train_scaled, dtype = "float32"),
  y = tf$convert_to_tensor(y_train, dtype = "float32"),
  epochs = as.integer(60),
  batch_size = as.integer(1024),
  validation_data = list(
    tf$convert_to_tensor(X_test_scaled, dtype = "float32"),
    tf$convert_to_tensor(y_test, dtype = "float32")
  ),
  class_weight = class_weights,
  callbacks = list(
    callback_early_stopping(
      monitor = "val_auc",
      patience = 15,
      restore_best_weights = TRUE,
      mode = "max"
    )
  ),
  verbose = 1
)

# ====================== PREDICTIONS ======================
pred_prob <- model$predict(
  tf$convert_to_tensor(X_test_scaled, dtype = "float32")
) %>% as.vector()

# Threshold tuning
thresholds <- seq(0.25, 0.60, by = 0.01)

results <- data.frame()

for(th in thresholds){
  pred_class <- ifelse(pred_prob > th, 1, 0)
  
  cm <- confusionMatrix(
    factor(pred_class, levels = c(0,1)),
    factor(testlocal$PitNextLap, levels = c(0,1)), 
    positive = "1"
  )
  
  results <- rbind(results, data.frame(
    threshold = th,
    recall    = cm$byClass["Recall"],
    precision = cm$byClass["Precision"],
    f1        = cm$byClass["F1"]
  ))
}

print(results[order(results$f1, decreasing = TRUE), ][1:8, ])

best_th <- results$threshold[which.max(results$f1)]
cat("\nBest threshold (by F1):", best_th, "\n")

cat_data_test <- as.data.frame(test[, cat_vars, drop = FALSE])
dummy_test <- predict(dummies, cat_data_test)
X_test <- cbind(as.data.frame(test[, num_cols, drop = FALSE]), dummy_test)
X_test_mat <- as.matrix(X_test)
pred_prob <- model$predict(
  tf$convert_to_tensor(X_test_mat, dtype = "float32")
) %>% as.vector()

pred_class <- ifelse(pred_prob > best_th, 1, 0)

NN_pred_class <- pred_class
submissionfileNN <- data.frame(id = test$id, PitNextLap = NN_pred_class)
write.csv(submissionfileNN, file = "F1SubmissionNNKeras.csv", row.names = FALSE)
```

This first attempt on the Neural Network resulted in an 87.57% prediction accuracy, good but worse than XGBoost.

Now its time to tune it.

```{r NN_Visualisation, echo = FALSe}
# 1. Training History (Learning Curves)
# Extract the metrics properly
h <- history$history   # this is usually a Python dict

# Convert to R data frame
epochs <- seq_along(h$loss)

history_df <- data.frame(
  epoch = epochs,
  loss = as.numeric(h$loss),
  val_loss = as.numeric(h$val_loss),
  auc = as.numeric(h$auc),
  val_auc = as.numeric(h$val_auc)
)

print(head(history_df))

# Plot Loss
plot(history_df$epoch, history_df$loss, type = "l", col = "blue", 
     xlab = "Epoch", ylab = "Loss", main = "Training History - Loss")
lines(history_df$epoch, history_df$val_loss, col = "red")
legend("topright", c("Train Loss", "Val Loss"), col = c("blue", "red"), lty = 1)

# Plot AUC
plot(history_df$epoch, history_df$auc, type = "l", col = "blue", 
     xlab = "Epoch", ylab = "AUC", main = "Training History - AUC")
lines(history_df$epoch, history_df$val_auc, col = "red")
legend("bottomright", c("Train AUC", "Val AUC"), col = c("blue", "red"), lty = 1)

# 2. Best Validation AUC
cat("Best Val AUC:", max(history$metrics$val_auc), "\n")

# 3. Confusion Matrix at best threshold
pred_prob <- model$predict(tf$convert_to_tensor(X_test_scaled, dtype = "float32")) %>% as.vector()
best_th <- 0.40   # start here, tune later

pred_class <- ifelse(pred_prob > best_th, 1, 0)
cm <- confusionMatrix(factor(pred_class, levels = c(0,1)),
                      factor(testlocal$PitNextLap, levels = c(0,1)), 
                      positive = "1")
print(cm)
```

As we can see from the Train and Validation AUC plot as well as the loss plot, 60 epochs was not enough for the model to converge, hence we will run it for 300 and add one extra layer at the start.

```{r NNTuned, echo = FALSE}
tf <- import("tensorflow")

# ====================== DATA PREPARATION ======================
X_train_scaled <- as.matrix(X_train)
X_test_scaled  <- as.matrix(X_test)

y_train <- as.matrix(as.factor(as.numeric(trainlocal$PitNextLap)))
y_test  <- as.matrix(as.factor(as.numeric(testlocal$PitNextLap)))

pos_weight <- sum(y_train == 0) / sum(y_train == 1)

cat("Positive weight:", round(pos_weight, 2), "\n")

# ====================== BUILD MODEL ======================
model <- keras_model_sequential()

model$add(layer_dense(units = 512, activation = "relu", 
                      input_shape = ncol(X_train_scaled)))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.4))

model$add(layer_dense(units = 256, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.4))

model$add(layer_dense(units = 128, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.35))

model$add(layer_dense(units = 64, activation = "relu"))
model$add(layer_batch_normalization())
model$add(layer_dropout(rate = 0.3))

model$add(layer_dense(units = 1, activation = "sigmoid"))

# ====================== COMPILE ======================
model$compile(
  optimizer = optimizer_adam(learning_rate = 0.0005),   # lower than before
  loss = "binary_crossentropy",
  metrics = c("accuracy", metric_auc(name = "auc"))
)

summary(model)
class_weights <- reticulate::dict("0" = 1.0, "1" = pos_weight * 1.3)
# ====================== TRAIN ======================
# ====================== IMPROVED TRAINING ======================

history <- model$fit(
  x = tf$convert_to_tensor(X_train_scaled, dtype = "float32"),
  y = tf$convert_to_tensor(y_train, dtype = "float32"),
  epochs = as.integer(300),           # Much higher
  batch_size = as.integer(1024),
  validation_data = list(
    tf$convert_to_tensor(X_test_scaled, dtype = "float32"),
    tf$convert_to_tensor(y_test, dtype = "float32")
  ),
  class_weight = class_weights,   # slight increase
  
  # === Key additions ===
  callbacks = list(
    callback_early_stopping(
      monitor = "val_auc",
      patience = 40,                    # more patient
      restore_best_weights = TRUE,
      mode = "max"
    ),
    # Reduce learning rate when progress slows
    callback_reduce_lr_on_plateau(
      monitor = "val_auc",
      factor = 0.5,
      patience = 15,
      min_lr = 0.00001,
      verbose = 1
    )
  ),
  verbose = 1
)

# ====================== PREDICTIONS ======================
pred_prob <- model$predict(
  tf$convert_to_tensor(X_test_scaled, dtype = "float32")
) %>% as.vector()

# Threshold tuning
thresholds <- seq(0.25, 0.60, by = 0.01)

results <- data.frame()

for(th in thresholds){
  pred_class <- ifelse(pred_prob > th, 1, 0)
  
  cm <- confusionMatrix(
    factor(pred_class, levels = c(0,1)),
    factor(testlocal$PitNextLap, levels = c(0,1)), 
    positive = "1"
  )
  
  results <- rbind(results, data.frame(
    threshold = th,
    recall    = cm$byClass["Recall"],
    precision = cm$byClass["Precision"],
    f1        = cm$byClass["F1"]
  ))
}

print(results[order(results$f1, decreasing = TRUE), ][1:8, ])

best_th <- results$threshold[which.max(results$f1)]
cat("\nBest threshold (by F1):", best_th, "\n")

cat_data_test <- as.data.frame(test[, cat_vars, drop = FALSE])
dummy_test <- predict(dummies, cat_data_test)
X_test <- cbind(as.data.frame(test[, num_cols, drop = FALSE]), dummy_test)
X_test_mat <- as.matrix(X_test)
pred_prob_NN <- model$predict(
  tf$convert_to_tensor(X_test_mat, dtype = "float32")
) %>% as.vector()

pred_class <- ifelse(pred_prob > 0.5, 1, 0)

submissionfileNN <- data.frame(id = test$id, PitNextLap = pred_prob_NN)
write.csv(submissionfileNN, file = "F1SubmissionNNKerasTuned.csv", row.names = FALSE)
```

Lets try an ensemble of XGBoost and Neural Network, as ensembles are usually better for prediction than individual models as they usually capture different information in which a combination of the two can be beneficial for predict

## Ensemble

```{r Ensembles, echo = FALSE}
pred_ensemble <- 0.65 * pred_prob_xg + 0.35 * pred_prob_NN   # tune these weights
pred_ensemble <- 0.5 * pred_prob_xg + 0.5 * pred_prob_NN 
pred_ensemble <- 0.35 * pred_prob_xg + 0.65 * pred_prob_NN 
pred_ensemble <- 0.55 * pred_prob_xg + 0.45 * pred_prob_NN 
pred_ensemble <- 0.6 * pred_prob_xg + 0.4 * pred_prob_NN 

xgPreddf <- read.csv("F1SubmissionXGTuned.csv")
RFPreddf <- read.csv("F1SubmissionRandomForest.csv")
NNPreddf <- read.csv("F1SubmissionNNKerasTuned.csv")

pred_xg_NN_RF <- 0.5*xgPreddf$PitNextLap + 0.25*NNPreddf$PitNextLap + 0.25*RFPreddf$PitNextLap


submissionfileEnsemble <- data.frame(id = test$id, PitNextLap = pred_xg_NN_RF)
write.csv(submissionfileEnsemble, file = "F1SubmissionEnsembleXG_RF_NN.csv", row.names = FALSE)


#Grok recommneded I try a different ratio based off of each model types individual strength
pred_xg_NN_RFGrok <- 0.5*xgPreddf$PitNextLap + 0.15*NNPreddf$PitNextLap + 0.35*RFPreddf$PitNextLap


submissionfileEnsemble <- data.frame(id = test$id, PitNextLap = pred_xg_NN_RFGrok)
write.csv(submissionfileEnsemble, file = "F1SubmissionEnsembleXG_RF_NN_Grok.csv", row.names = FALSE)

predict_reg2

#Final ensemble with logistic classification included as well

pred_xg_NN_RF_LogGrok <- 0.5*xgPreddf$PitNextLap + 0.1*NNPreddf$PitNextLap + 0.3*RFPreddf$PitNextLap + 0.1*predict_reg2


submissionfileEnsemble <- data.frame(id = test$id, PitNextLap = pred_xg_NN_RF_LogGrok)
write.csv(submissionfileEnsemble, file = "F1SubmissionEnsembleXG_RF_NN_Log.csv", row.names = FALSE)
```


Prediction was 94.62% on the kaggle test set, our best AUC so far and we will leave it at that. Currently within 1% of the number one spot and to avoid overfitting to the test set we will leave it as is.



```{r eval=FALSE, EnsembleTune, echo=FALSE, include=FALSE}
traindata <- train[train_idx,-c(1,2)]
testdata<- train[-train_idx,-c(1,2)]
#traindata$PitNextLap <- as.numeric(traindata$PitNextLap) - 1 # Important: 0 and 1
#testdata$PitNextLap  <- as.numeric(testdata$PitNextLap) - 1

set.seed(123) # For reproducibility
k <- 5
data <- traindata

# 2. Create Fold Indices
folds <- cut(seq(1, nrow(data)), breaks = k, labels = FALSE)
errors <- numeric(k)
thresholds <- seq(0.25, 0.55, by = 0.05)
prediction_mat <- matrix(0, nrow = k, ncol = length(thresholds)*2)

ensemble_ratios <- c(0.35,0.4,0.45,0.5)
# 3. Perform Cross-Validation
for(i in 1:k) {
    # Split into train and test
    test_indices <- which(folds == i, arr.ind = TRUE)
    testlocal <- data[test_indices, ]
    trainlocal <- data[-test_indices, ]
    
    # Identify categorical variables
    cat_vars <- names(trainlocal)[sapply(trainlocal, function(x) is.factor(x) || is.character(x))]
    
    # ====================== FEATURE PREPARATION ======================
    if(length(cat_vars) > 0){
      
      # Train dummies
      cat_data_train <- as.data.frame(trainlocal[, cat_vars, drop = FALSE])
      dummies <- dummyVars(~ ., data = cat_data_train)
      dummy_train <- predict(dummies, cat_data_train)
      
      # Numeric columns
      num_cols <- setdiff(names(trainlocal), c(cat_vars, "PitNextLap"))
      X_train <- cbind(as.data.frame(trainlocal[, num_cols, drop = FALSE]), dummy_train)
      
      # Test dummies (using same dummy structure)
      cat_data_test <- as.data.frame(testlocal[, cat_vars, drop = FALSE])
      dummy_test <- predict(dummies, cat_data_test)
      X_test <- cbind(as.data.frame(testlocal[, num_cols, drop = FALSE]), dummy_test)
      
    } else {
      X_train <- as.data.frame(trainlocal %>% select(-PitNextLap))
      X_test  <- as.data.frame(testlocal %>% select(-PitNextLap))
    }
    # ====================== FIX FEATURE ALIGNMENT ======================
    
    # Make sure test has exactly the same columns as train (in same order)
    common_cols <- intersect(names(X_train), names(X_test))
    
    # Add missing columns to test (filled with 0)
    missing_in_test <- setdiff(names(X_train), names(X_test))
    for (col in missing_in_test) {
      X_test[[col]] <- 0
    }
    
    # Add missing columns to train (just in case)
    missing_in_train <- setdiff(names(X_test), names(X_train))
    for (col in missing_in_train) {
      X_train[[col]] <- 0
    }
    
    # Re-order test columns to match train exactly
    X_test <- X_test[, names(X_train), drop = FALSE]
    
    # Now create DMatrix again
    dtrain <- xgb.DMatrix(data = as.matrix(X_train), label = trainlocal$PitNextLap)
    dtest  <- xgb.DMatrix(data = as.matrix(X_test),  label = testlocal$PitNextLap)
    
    cat("Feature alignment fixed. Total features:", ncol(X_train), "\n")
    
    # ====================== CLASS IMBALANCE ======================
    pos_weight <- sum(trainlocal$PitNextLap == 0) / sum(trainlocal$PitNextLap == 1)
    cat("scale_pos_weight:", round(pos_weight, 2), "\n")
    
    # ====================== CHECK CLASS DISTRIBUTION ======================
    cat("Train class distribution:\n")
    print(table(trainlocal$PitNextLap))
    cat("Test class distribution:\n")
    print(table(testlocal$PitNextLap))
    
    
    # ====================== XGBoost MODEL TRAINING ======================
    params <- list(
      objective         = "binary:logistic",
      eval_metric       = "auc",          # changed from logloss
      eta               = 0.03,           # decreased from 0.05 for a slower learning rate
      max_depth         = 7,              # increased from 6
      subsample         = 0.85,           # increased from 0.8
      colsample_bytree  = 0.75,           # increased from 0.8
      scale_pos_weight  = pos_weight, 
      gamma             = 0.5,            
      min_child_weight  = 3,
      max_delta_step    = 1,
      nthread           = parallel::detectCores() - 1
    )
    
    set.seed(123)
    
    xgb_model <- xgb.train(
      params            = params,
      data              = dtrain,
      nrounds           = 500,
      watchlist         = list(train = dtrain, test = dtest),
      early_stopping_rounds = 80,
      verbose           = 20
    )
    
    # ====================== XGBoost PREDICTIONS ======================
    pred_probXG <- predict(xgb_model, dtest)
    
    # ====================== Neural Net Model Training ======================
    X_train_scaled <- as.matrix(X_train)
    X_test_scaled  <- as.matrix(X_test)
    
    y_train <- as.matrix(as.factor(trainlocal$PitNextLap))
    y_test  <- as.matrix(as.factor(testlocal$PitNextLap))
    
    pos_weight <- sum(y_train == 0) / sum(y_train == 1)
    
    cat("Positive weight:", round(pos_weight, 2), "\n")
    
    # ====================== BUILD MODEL ======================
    model <- keras_model_sequential()
    
    model$add(layer_dense(units = 512, activation = "relu", 
                          input_shape = ncol(X_train_scaled)))
    model$add(layer_batch_normalization())
    model$add(layer_dropout(rate = 0.4))
    
    model$add(layer_dense(units = 256, activation = "relu"))
    model$add(layer_batch_normalization())
    model$add(layer_dropout(rate = 0.4))
    
    model$add(layer_dense(units = 128, activation = "relu"))
    model$add(layer_batch_normalization())
    model$add(layer_dropout(rate = 0.35))
    
    model$add(layer_dense(units = 64, activation = "relu"))
    model$add(layer_batch_normalization())
    model$add(layer_dropout(rate = 0.3))
    
    model$add(layer_dense(units = 1, activation = "sigmoid"))
    
    # ====================== COMPILE ======================
    model$compile(
      optimizer = optimizer_adam(learning_rate = 0.0005),   # lower than before
      loss = "binary_crossentropy",
      metrics = c("accuracy", metric_auc(name = "auc"))
    )
    
    class_weights <- reticulate::dict("0" = 1.0, "1" = pos_weight)
    # ====================== TRAIN ======================
    # ====================== IMPROVED TRAINING ======================
    
    history <- model$fit(
      x = tf$convert_to_tensor(X_train_scaled, dtype = "float32"),
      y = tf$convert_to_tensor(y_train, dtype = "float32"),
      epochs = as.integer(300),           # Much higher
      batch_size = as.integer(1024),
      validation_data = list(
        tf$convert_to_tensor(X_test_scaled, dtype = "float32"),
        tf$convert_to_tensor(y_test, dtype = "float32")
      ),
      class_weight = class_weights,   # slight increase
      
      # === Key additions ===
      callbacks = list(
        callback_early_stopping(
          monitor = "val_auc",
          patience = 40,                    # more patient
          restore_best_weights = TRUE,
          mode = "max"
        ),
        # Reduce learning rate when progress slows
        callback_reduce_lr_on_plateau(
          monitor = "val_auc",
          factor = 0.5,
          patience = 15,
          min_lr = 0.00001,
          verbose = 1
        )
      ),
      verbose = 1
    )
    
    # ====================== PREDICTIONS ======================
    pred_probNN <- model$predict(
      tf$convert_to_tensor(X_test_scaled, dtype = "float32")
    ) %>% as.vector()
    
    
    
    # Testing different ensemble ratios
    pred_XGVvals <- pred_probXG*ensemble_ratios + pred_probNN*(1-ensemble_ratios)
    pred_NNVvals <- pred_probXG*(1-ensemble_ratios) + pred_probNN*ensemble_ratios
    pred_prob_ensemble <- c(pred_XGVvals, pred_NNVvals)
    ratio_names <- c(ensemble_ratios, (1-ensemble_ratios), (1-ensemble_ratios), ensemble_ratios)
    


    # Try different thresholds
    pred_num <- 1
    len_ens <- length(ensemble_ratios)
    pred_mat <- matrix(0, nrow = k, ncol = length(ratio_names)*length(thresholds)*2)
    matcol <- 1
    for(pred in pred_prob_ensemble){
      pred_comb <- c(ratio_names[pred_num],ratio_names[pred_num+len_ens])
      for(th in thresholds){
        pred_class <- ifelse(pred > th, 1, 0)
        accuracy_score <- sum(testlocal$PitNextLap == pred_class)/nrow(testlocal)
        true_positives <- testlocal$PitNextLap[which(testlocal$PitNextLap == 1)]
        recall <- sum(true_positives == pred_class[which(testlocal$PitNextLap == 1)])/length(true_positives)
        prediction_mat[i, thcol] <- accuracy_score
        prediction_mat[i, thcol+1] <- recall
        cat("Accuracy Score of:", accuracy_score, "threshold:",th, "ratio:", pred_comb, "\n")
        cat("Recall Score of:", recall, "\n")
        matcol <- matcol + 2
      }
      pred_num <- pred_num + 1
    }

}

```